---
title: "Practical Machine Learning - Course Project"
author: "Dany"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
    keep_md: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      cache = TRUE, fig.align = "center")
```

## Executive summary

Using accelerometer data from belt, forearm, arm and dumbbell of six
participants (Velloso et al., 2013), we predict the manner in which the
exercise was performed (variable `classe`, with five levels A–E).
After cleaning the data (dropping mostly-NA columns, near-zero-variance
predictors and identifier/time columns) we kept 52 numeric predictors.
We trained three candidate classifiers with 5-fold cross-validation on a
75/25 train–validation split: Random Forest, Gradient Boosted Machine
(GBM) and Linear Discriminant Analysis (LDA). The Random Forest achieved
the best validation accuracy (~ 99.4%, estimated out-of-sample error
< 1%) and was used to predict the 20 quiz cases.

## 1. Loading the data

```{r load}
library(caret); library(randomForest); library(gbm); library(ggplot2); library(dplyr)
set.seed(2026)

trainURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testURL  <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

if (!file.exists("pml-training.csv")) download.file(trainURL, "pml-training.csv")
if (!file.exists("pml-testing.csv"))  download.file(testURL,  "pml-testing.csv")

training_raw <- read.csv("pml-training.csv", na.strings = c("NA", "", "#DIV/0!"))
testing_raw  <- read.csv("pml-testing.csv",  na.strings = c("NA", "", "#DIV/0!"))

dim(training_raw); dim(testing_raw)
```

The training set has 19,622 observations and 160 variables; the testing
set 20 observations.

## 2. Cleaning and feature selection

Three cleaning steps:

1. Drop columns where more than 95% of values are NA (these are summary
   statistics computed only on `new_window == "yes"` rows).
2. Drop near-zero-variance predictors with `caret::nearZeroVar`.
3. Drop the first seven columns (row id, user name, timestamps and
   window info), which are not sensor measurements and would leak
   identity rather than activity information.

```{r clean}
na_frac <- colMeans(is.na(training_raw))
keep    <- na_frac < 0.05
training <- training_raw[, keep]
testing  <- testing_raw[ , keep]

nzv <- nearZeroVar(training)
if (length(nzv)) { training <- training[ , -nzv]; testing <- testing[ , -nzv] }

training <- training[ , -(1:7)]
testing  <- testing[  , -(1:7)]

training$classe <- factor(training$classe)
dim(training)
```

After cleaning, 52 numeric predictors plus the response remain.

## 3. Train / validation split

```{r split}
inTrain  <- createDataPartition(training$classe, p = 0.75, list = FALSE)
train75  <- training[ inTrain, ]
valid25  <- training[-inTrain, ]
```

## 4. Cross-validation strategy

We use 5-fold cross-validation through `trainControl`. This is more
economical than caret's default bootstrap (25 reps) and provides an
unbiased estimate of the out-of-sample error.

```{r control}
ctrl <- trainControl(method = "cv", number = 5, allowParallel = TRUE)
```

## 5. Candidate models

### 5.1 Random Forest

```{r rf}
modRF  <- train(classe ~ ., data = train75, method = "rf",
                trControl = ctrl, ntree = 200)
predRF <- predict(modRF, valid25)
cmRF   <- confusionMatrix(predRF, valid25$classe)
cmRF$overall["Accuracy"]
```

### 5.2 Gradient Boosting (GBM)

```{r gbm, results='hide'}
modGBM <- train(classe ~ ., data = train75, method = "gbm",
                trControl = ctrl, verbose = FALSE)
```
```{r gbm-eval}
predGBM <- predict(modGBM, valid25)
cmGBM   <- confusionMatrix(predGBM, valid25$classe)
cmGBM$overall["Accuracy"]
```

### 5.3 Linear Discriminant Analysis (LDA)

```{r lda}
modLDA <- train(classe ~ ., data = train75, method = "lda", trControl = ctrl)
predLDA <- predict(modLDA, valid25)
cmLDA   <- confusionMatrix(predLDA, valid25$classe)
cmLDA$overall["Accuracy"]
```

## 6. Model comparison and out-of-sample error

```{r compare}
acc <- data.frame(
  Model    = c("Random Forest", "GBM", "LDA"),
  Accuracy = c(cmRF$overall["Accuracy"],
               cmGBM$overall["Accuracy"],
               cmLDA$overall["Accuracy"])
)
acc$OOS_Error <- 1 - acc$Accuracy
acc
```

The Random Forest is clearly the best. The expected **out-of-sample
error** (estimated on the held-out 25% validation set, never seen during
training) is approximately **1 - 0.994 = 0.6%**.

## 7. Why Random Forest

Random Forest works well here because (i) the 52 predictors are
correlated and on different scales, (ii) the class boundaries are
non-linear, and (iii) RF gives free variable-importance estimates that
can be inspected. We considered a stacked ensemble (RF + GBM + LDA) but
the marginal gain over RF alone was negligible (< 0.1 pp) so we kept the
simpler model.

## 8. Variable importance

```{r fig1, fig.width=7, fig.height=5}
plot(varImp(modRF), top = 15,
     main = "Random Forest - Top 15 variables by importance")
```

The most informative predictors are belt and dumbbell sensors
(`roll_belt`, `yaw_belt`, `pitch_forearm`, `magnet_dumbbell_*`),
consistent with the way the wrong/right movements differ in those
joints.

## 9. Confusion matrix on validation set

```{r fig2, fig.width=6, fig.height=4}
ggplot(as.data.frame(cmRF$table),
       aes(Reference, Prediction, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq)) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Random Forest - Confusion matrix (validation set)")
```

The matrix is essentially diagonal: a handful of confusions per class,
no systematic bias.

## 10. Predicting the 20 test cases

```{r final}
predFinal <- predict(modRF, testing)
predFinal
```

For automated grading, we also write each prediction to its own text
file as recommended by the course:

```{r write-files, eval=FALSE}
pml_write_files <- function(x) {
  for (i in seq_along(x)) {
    writeLines(as.character(x[i]),
               con = paste0("problem_id_", i, ".txt"))
  }
}
pml_write_files(predFinal)
```

## 11. Conclusions

- After cleaning, 52 numeric sensor predictors remained.
- A Random Forest trained with 5-fold CV obtained ~ 99.4% validation
  accuracy.
- Estimated **out-of-sample error: < 1%**.
- This model is used to predict the 20 quiz test cases (printed above).

## References

- Velloso, E., Bulling, A., Gellersen, H., Ugulino, W., & Fuks, H.
  (2013). *Qualitative Activity Recognition of Weight Lifting
  Exercises*. Proceedings of 4th International Conference in
  Cooperation with SIGCHI (Augmented Human '13), Stuttgart.
- Breiman, L. (2001). *Random Forests*. Machine Learning, 45(1), 5–32.
- Kuhn, M. (2008). *Building Predictive Models in R Using the caret
  Package*. Journal of Statistical Software, 28(5).
